---
title: "An Introduction to BFI"
author: 
- Hassan Pazira
- Marianne Jonker
date: "`r Sys.Date()`" # "`r format(Sys.time(), '%B %d, %Y')`"
output:
  #rmarkdown::html_vignette
  knitr:::html_vignette:
    toc: yes
    #toc_depth: 3
    #theme: bootstrap
# output:
#   html_document:
#     fig_caption: yes
#     toc: yes
#     toc_depth: 3
#     theme: lumen
description: |
  How to use the BFI package.
vignette: >
  %\VignetteIndexEntry{An Introduction to BFI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
require(knitr)
opts_chunk$set(
  collapse = F # T for red
)
```


## Overview

The R package `BFI` (**B**ayesian **F**ederated **I**nference) provides several functions to carry out the Bayesian Federated Inference method for two types of models (`GLM` and `Survival`) using multicenter data without the need to combine or share them. This tutorial focuses on `GLM` models. Two commonly used families for `GLM` models, `"binomial"` and `"gaussian"`, are available for this version of the package.
The mostly using functions include `bfi()`, `MAP.estimation()`, and `inv.prior.cov()`.
In the following, we will see how the `BFI` package can be applied to real datasets included in the package.


## How to use it?

Before we go on, we first install and load the `BFI` package:

```{r , include = FALSE}
# First set a CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com/"))
```

```{r , message=FALSE, results='hide'}
# Install and load the BFI package from CRAN:
install.packages("BFI")
library(BFI)
```
By using the following code, we can see that there are two available datasets in the package: `trauma` and `Nurses`.
```{r}
data(package = "BFI")
```

The `trauma` dataset can be utilized for the `"binomial"` family and `Nurses` dataset can be used for `"gaussian"` family. To avoid repetition, we will only use the `trauma` dataset.
By loading the package, the datasets included will be loaded and can be inspected as follows:
```{r}
# Get the number of rows and columns
dim(trauma)

# To get an idea of the dataset, print the first 7 rows
head(trauma, 7)
```

This dataset consists of data of 371 trauma patients from three hospitals (peripheral hospital without a neuro-surgical unit, `'status=1'`, peripheral hospital with a neuro-surgical unit, `status=2`, and academic medical center, `status=3`).

As we can see, it has 6 columns:
```{r}
(col_name <- colnames(trauma))
```
The covariates `sex` (dichotomous), `age` (continuous), `ISS` (Injury Severity Score, continuous), and `GCS` (Glasgow Coma Scale, continuous) are the predictors, and `mortality` is the response variable. `hospital` is a categorical variable which indicates the hospitals involved in the study.
For more information about this dataset use 
```{r}
# Get some info about the dataset from the help file
?trauma
```

We will analyse the data with a `logistic` regression model. 
First we standardize the covariates. This is not necessary for the analysis, but is done for the interpretability of the accuracy of the estimates.
```{r}
trauma$age <- scale(trauma$age)
trauma$ISS <- scale(trauma$ISS) 
trauma$GCS <- scale(trauma$GCS) 
trauma$hospital <- as.factor(trauma$hospital)
```

By using the following code we can see there are three hospitals involved in the study:
```{r}
length(levels(trauma$hospital))
```

### MAP estimations

Therefore, the `MAP.estimation` function should be applied to these 3 local datasets separately to obtain the MAP estimations. Note that, in practice, we do not have access to the combined data, and each center should perform the analysis independently and send the output to the central server, as follows:
```{r}
# Center 1:
X1 <- data.frame(sex=trauma$sex[trauma$hospital==1],
                 age=trauma$age[trauma$hospital==1],
                 ISS=trauma$ISS[trauma$hospital==1],
                 GCS=trauma$GCS[trauma$hospital==1])
Lambda1 <- inv.prior.cov(X1, lambda=0.01, L=3, family="binomial")
fit1 <- MAP.estimation(y=trauma$mortality[trauma$hospital==1], X=X1, family="binomial", Lambda=Lambda1)
summary(fit1)

# Center 2:
X2 <- data.frame(sex=trauma$sex[trauma$hospital==2],
                 age=trauma$age[trauma$hospital==2],
                 ISS=trauma$ISS[trauma$hospital==2],
                 GCS=trauma$GCS[trauma$hospital==2])
Lambda2 <- inv.prior.cov(X2, lambda=0.01, L=3, family="binomial")
fit2 <- MAP.estimation(y=trauma$mortality[trauma$hospital==2], X=X2, family="binomial", Lambda=Lambda2)
summary(fit2)

# Center 3:
X3 <- data.frame(sex=trauma$sex[trauma$hospital==3],
                 age=trauma$age[trauma$hospital==3],
                 ISS=trauma$ISS[trauma$hospital==3],
                 GCS=trauma$GCS[trauma$hospital==3])
Lambda3 <- inv.prior.cov(X3, lambda=0.01, L=3, family="binomial")
fit3 <- MAP.estimation(y=trauma$mortality[trauma$hospital==3], X=X3, family="binomial", Lambda=Lambda3)
summary(fit3)
```
It can be seen that all algorithms converged (`Convergence: 0`).
To see more information about the dataset, such as the number of observations and parameters, we can use the output of the `MAP.estimation` function as follows:
```{r}
# number of samples in center 1
fit1$n
# number of parameters in center 1
fit1$np

# number of samples in center 2
fit2$n

# number of samples in center 3
fit3$n
```
Additionally, before conducting the analysis, we can use the `n.par` function to retrieve this information.

### BFI estimations

Then, only the outputs from the local centers (i.e., `fit1`,`fit2`, and `fit3`) should be sent to the central server for further analysis.
On the central server, the `bfi()` function can be used to obtain the BFI estimations:
```{r}
theta_hats <- list(fit1$theta_hat, fit2$theta_hat, fit3$theta_hat)
A_hats     <- list(fit1$A_hat, fit2$A_hat, fit3$A_hat)
Lambda_com <- inv.prior.cov(X1, lambda=0.01, L=3, family="binomial")
Lambdas    <- list(Lambda1, Lambda2, Lambda3, Lambda_com)
BFI_fits   <- bfi(theta_hats, A_hats, Lambdas)
summary(BFI_fits, cur_mat=TRUE)
```

### Comparison

To compare the performance of the BFI methodology, we can combine the datasets and obtain the MAP estimations based on the combined data:

```{r}
# MAP estimates of the combined data:
X_combined  <- data.frame(sex=trauma$sex,
                          age=trauma$age,
                          ISS=trauma$ISS,
                          GCS=trauma$GCS)
Lambda   <- inv.prior.cov(X=X_combined, lambda=0.1, L=3, family="binomial")
fit_comb  <- MAP.estimation(y=trauma$mortality, X=X_combined, family="binomial", Lambda=Lambda) 
summary(fit_comb, cur_mat=TRUE)
```

Now, we can see the difference between the BFI and combined estimates:
```{r}
# Squared Errors:
(fit_comb$theta_hat - BFI_fits$theta_hat)^2
```
whicha are close to zero, as expected!

For more details see the following references.

## References
Jonker M.A., Pazira H. and Coolen A.C.C. (2024). Bayesian federated inference for estimating statistical models based on non-shared multicenter data sets, Statistics in Medicine, 43(12): 2421-2438. <https://doi.org/10.1002/sim.10072>

Pazira H., Massa E., Weijers J.A.M., Coolen A.C.C. and Jonker M.A. (2024). Bayesian Federated Inference for Survival Models, arXiv. <https://arxiv.org/abs/2404.17464>

Jonker M.A., Pazira H. and Coolen A.C.C. (2024b). Bayesian Federated Inference for regression models with heterogeneous multi-center populations, arXiv. <https://arxiv.org/abs/2402.02898>



## Contact

If you find any errors, have any suggestions, or would like to request that something be added, please file an issue at [issue report](https://github.com/hassanpazira/BFI/issues/) or send an email to: hassan.pazira@radboudumc.nl or Marianne.Jonker@radboudumc.nl.
