---
title: "An Introduction to BFI"
author: 
- Hassan Pazira
- Marianne Jonker
date: "`r Sys.Date()`" # "`r format(Sys.time(), '%B %d, %Y')`"
output:
  #rmarkdown::html_vignette
  knitr:::html_vignette:
    toc: yes
    #toc_depth: 3
    #theme: bootstrap
# output:
#   html_document:
#     fig_caption: yes
#     toc: yes
#     toc_depth: 3
#     theme: lumen
description: |
  How to use the BFI package.
vignette: >
  %\VignetteIndexEntry{An Introduction to BFI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
require(knitr)
opts_chunk$set(
  collapse = F # T for red
)
```


## Overview

The R package `BFI` (**B**ayesian **F**ederated **I**nference) provides several functions to carry out the Bayesian Federated Inference method for two kinds of models (`GLM` and `Survival`) with multicenteral data without combining/sharing them. In this tutorial we focus on `GLM` only, so that this version of the package is available for two commonly used families: `"binomial"` and `"gaussian"`.
The mostly using functions include `bfi()`, `MAP.estimation()`, and `inv.prior.cov()`.
In the following, we will see how the `BFI` package can be applied to real datasets included in the package.


## How to use it?

Before we go on, we first install and load the `BFI` package using the `devtools` package:

```{r , message=FALSE}
# First install and load the package 'devtools'
if(!require(devtools)) {install.packages("devtools")}
library(devtools)

# Now install BFI from GitHub
#devtools::install_github("hassanpazira/BFI", force = TRUE)

# load BFI
library(BFI)
```

By the following code we can see there two available datasets in the package: `trauma` and `Nurses`.
```{r}
data(package = "BFI")
```

The `trauma` data can be utilized for the `"binomial"` family and `Nurses` data can be used for `"gaussian"`. To avoid repetition, we only use the `trauma` data set.
Load and inspect the `trauma` data as follows:
```{r}
# Load 'trauma' in the R workspace
data("trauma")

# Get the number of rows and columns
dim(trauma)

# To get an idea of the dataset, print the first 7 rows
head(trauma, 7)
```

This data set consists of data of 371 trauma patients from three hospitals (peripheral hospital without a neuro-surgical unit, status=1, peripheral hospital with a neuro-surgical unit, status=2, and academic medical center, status=3).

As we can see it has 6 columns:
```{r}
(col_name <- colnames(trauma))
```
The covariates `sex` (dichotomous), `age` (continuous), `ISS` (Injury Severity Score, continuous), and `GCS` (Glasgow Coma Scale, continuous) are the predictors, and `mortality` is the response variable. `hospital` is a categorical variable which indicates the hospitals involved in the study.
For more information about this dataset use 
```{r}
# Get some info about the dataset from the help file
?trauma
```

```{r}
```





## References

