---
title: "BFI in Python"
# author: 
# - Hassan Pazira
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  #rmarkdown::html_vignette
  knitr:::html_vignette:
    toc: yes
# output:
#   pdf_document:
#     fig_caption: yes
#     toc: yes
#     toc_depth: 3
description: |
  How to use the BFI package in Python environments.
vignette: >
  %\VignetteIndexEntry{BFI in Python}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
require(knitr)
opts_chunk$set(
  collapse = F # T for red
)
```


## Overview

The R package `BFI` (**B**ayesian **F**ederated **I**nference) 


## Python Instalation

The steps to install the latest version of Python on 'MacOS' and 'Ubuntu':

### For MacOS

```{r setup, message=FALSE}
# install BFI from GitHub
devtools::install_github("hassanpazira/BFI")
# load BFI
library(BFI)

# load `trauma` data
data("trauma")
# attach the data
attach(trauma)
```

### For Ubuntu

```{r}
?trauma
```

## Installing PIP

PIP (a Python package manager) helps us to install and use various packages/modules in Python programming.

Install/Upgrade the latest version of PIP by running the following command:

### On MacOS

```{r}
colnames(trauma)
```

### On Ubuntu

```{r}
colnames(trauma)
```

## Instalation of required moduls through PIP


### Installing the 'pandas' module


### Installing the 'numpy' module


### Installing the 'rpy2' module


## Python script

Now move on to work with 'rpy2' inside a Python script!
First type the following command in the terminal window to go to the Python Environment:

### Importing required packages and subpackages


### Installation of R packages from CRAN and GitHub:

### Importing the required functions


### Installing required R packages from CRAN


### Installing the 'BFI' package from GitHub


## Examples

### Data Simulation for center 1


### Data Simulation for center 2


### MAP estimates at center 1

### MAP estimates at center 2

### BFI at central center




## References

